---
title: "drhur-vignette"
author: "Yufei Sun"
date: "9/15/2020"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{drhur-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include=FALSE}
library(kableExtra)
```

# Overview
**drhur** is a function that calls a run_tutorial function to open a `learnr` project. The project list is continually updated along with the R workshop with Dr. Hu.

# Core Functions
**drhur** mainly includes the following parts：

```{r, results='asis', echo=FALSE}
txt <- data.frame(
  Function = 
    c('<a href="#playing">playing</a>',
      '<a href="#munging">munging</a>',
      '<a href="#analyzing">analyzing</a>',
      '<a href="#visualizing">visualizing</a>&nbsp;&nbsp;'),
  Description = 
    c('**Playing around with R:** Intro to basic data structure in R; inputting, outputting, and creating data.',
      paste('**Data munging: ** Generalizing (such as sorting and summarizing) and extracting (selecting or adjusting certain columns or rows) data frames.'),
      paste('**Data analyzing:** Analyze data with descriptive and multivariate tools and diagnoses.'),
      paste("**Data presenting:** Present descriptive and analytic results with 'ggplot2'")),
  Usage = 
    c('`drhur(playing)`',
      paste('`drhur(munging)`'),
      paste('`drhur(analyzing)`'),
      paste('`drhur(visualizing)`'))
  )

kable(txt, format = "html", escape = FALSE, align = c('l', 'l')) %>%
  kable_styling(full_width = FALSE, position = "left") %>%
  column_spec(1, bold = TRUE, monospace = TRUE, border_right = TRUE) %>%
  column_spec(2, width = "28em") %>%
  column_spec(3, width = "28em")
```

# Conclusion
It is a work in progress and feedback is always welcome. Please open an issue on [GitHub](https://github.com/sammo3182/drhur) if you find a bug or wish to submit a feature request.

Check out the [GitHub project’s page](GitHub project’s page); from there you can see the latest updates and also submit feature requests.