---
title: "drhur: Learn R with Dr. Hu Easily and Happily"
author: "Yue Hu and Yufei Sun"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{drhur: Learn R with Dr. Hu Easily and Happily}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE
)
```

## Overview

Learning a programming language such as R can be (rewardingly) challenging.
`drhur` is a pedagogical software to make the learning process easier and more enjoyable.
The package includes tutorials on basic data science skills with R, such as basic R syntax, data import/export, munging, and visualization.
The goal of this package is to provide an interactive platform for learners to easily and efficiently enter the R world.

The ideal way to acquire these skills is through the "Learning R with Dr. Hu" workshop, conducted by Associate Professor Yue Hu from the Department of Political Science at Tsinghua University. 
The modules included in this package correspond to the content of the initial chapters of this workshop series and Dr. Hu's forthcoming textbook on the R programming language.
The workshop series and the book aim to cover knowledge and skills about R, ranging from basic to advanced techniques, such as spatial analysis, network analysis, and text analysis.
You can find more relevant information on the [author's website](https://sammo3182.github.io/software/drhur/) or by [contacting him](mailto:yuehu@tsinghua.edu.cn) directly.

## Usage

This package is constructed using the `learnr` function, which launches a Shiny app allowing learners to interact seamlessly with instructions and tasks. 
One can conveniently load the package and access the tutorials through the `drhur()` function. 
For example, the visualization tutorial can be initiated with the following commands:

```{r eval = FALSE}
if (!require(drhur)) install.packages("drhur")
library(drhur)

drhur("visualize")
```


## Conclusion

`drhur` aims to help R learners to get in touch with the most common applied skills of the programming language. 
We encourage users to consult the help files for more details. 

The development of the package is ongoing. 
Please contact us with any questions, bug reports, and comments. Free to open [an issue on the software's GitHub page](https://github.com/sammo3182/drhur/issues) if you have any question or a feature request.

## Affiliation

Yue Hu

Political Science,    
Tsinghua Univeristy   
Email: <yuehu@tsinghua.edu.cn>    
Website: <https://sammo3182@tsinghua.edu.cn>
