---
title: "drhur: Learn R with Dr. Hu Easily and Happily"
author: "Yue Hu and Yufei Sun"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{drhur: Learn R with Dr. Hu Easily and Happily}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE
)
```

## Overview

Learning a programming language such as R can be (rewardingly) challenging.
`drhur` is a pedagogical software to make the learning process easier and more enjoyable.
The package includes tutorials on basic data science skills with R, such as basic R syntax, data import/export, munging, and visualization.
The goal of this package is to provide an interactive platform for learners to easily and efficiently enter the R world.

The ideal way to acquire these skills is through the "Learning R with Dr. Hu" workshop, conducted by Associate Professor Yue Hu from the Department of Political Science at Tsinghua University. 
The modules included in this package correspond to the content of the initial chapters of this workshop series and Dr. Hu's forthcoming textbook on the R programming language.
The workshop series and the book aim to cover knowledge and skills about R, ranging from basic to advanced techniques, such as spatial analysis, network analysis, and text analysis.
You can find more relevant information on the [author's website](https://www.drhuyue.site/software/drhur/) or by [contacting him](mailto:yuehu@tsinghua.edu.cn) directly.

## Usage

Each workshop is a self-contained [Quarto Live](https://r-wasm.github.io/quarto-live/) document whose code cells run directly in the browser via WebAssembly (`webR`), so learners can read the instructions and run the exercises side by side without setting up a local R session.
One can conveniently load the package and open a workshop through the `drhur()` function, which renders the document and opens it in the browser.
For example, the data-visualization workshop can be initiated with the following commands:

```{r eval = FALSE}
if (!require(drhur)) install.packages("drhur")
library(drhur)

# Open as a scrollable HTML live document (the default)
drhur("visual_basic")

# Or open the same workshop as a reveal.js slide deck
drhur("visual_basic", style = "slide")
```

The *Basics* series currently includes six workshops: `algebra`, `play`, `munging`, `descriptive`, `visual_basic`, and `loop`. Each is available in two language editions, selected with the `lang` argument---simplified Chinese (`"cn"`, the default) and English (`"en"`):

```{r eval = FALSE}
# Open the visualization workshop in English
drhur("visual_basic", lang = "en")
```

Rendering requires the [Quarto](https://quarto.org) command-line tool to be installed, and running the interactive code cells requires an internet connection (the `webR` runtime and the workshop's R packages are downloaded in the browser at view time).


## Conclusion

`drhur` aims to help R learners to get in touch with the most common applied skills of the programming language. 
We encourage users to consult the help files for more details. 

The development of the package is ongoing. 
Please contact us with any questions, bug reports, and comments. Free to open [an issue on the software's GitHub page](https://github.com/sammo3182/drhur/issues) if you have any question or a feature request.

## Affiliation

Yue Hu

Political Science,    
Tsinghua Univeristy   
Email: <yuehu@tsinghua.edu.cn>    
Website: <https://www.drhuyue.site>
