---
title: "Data Wrangling"
subtitle: "Learn R with Dr. Hu and His Friends"
author: "Yue Hu (Department of Political Science, Tsinghua University)"
output:
  learnr::tutorial:
    language: "en"
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      out.width = "100%")

if (!require(pacman)) install.packages("pacman")

pacman::p_load(drhur,
               here,
               rio,
               tidyverse)

if (!require("dplyr")) {
  install.packages("dplyr")
  library("dplyr")
} else {
  library("dplyr")
}
```


## Key Points

- **Research questions**
    - We have several survey datasets — Afrobarometer, Asianbarometer, Latinobarómetro — and wish to combine them into a single global barometer dataset. How should they be merged? *(Hint: row-binding.)*
    - We hold (1) annual survey data on inequality perception across Chinese provinces and (2) annual data on within-province income gaps. How should the two be merged to study the effect of within-province inequality on residents' inequality perception? *(Hint: index-based merging on province and year.)*

- Single-dataset operations
    - Exploring
    - Selecting
    - Filtering
    - Modifying
- Cross-dataset operations
    - Direct merging
    - Index-based merging


## The Demonstration Dataset

This chapter uses a sample drawn from Wave 7 of the [World Values Survey](https://www.worldvaluessurvey.org/wvs.jsp), originally founded by the late political scientist Ronald Inglehart. The sample comprises 2 % of the WVS7 records and includes 24 variables. For variable definitions, run `?drhur::wvs7`.


## Data Exploration

Data exploration is the preliminary inspection of an unfamiliar dataset's composition, structure, format, and content. It is the first — and arguably the most important — step in any analysis.


### A Glimpse of the Raw Data

```{r glimpse, exercise = TRUE, exercise.eval = TRUE}
wvs7
```

In line with the **tidy data** convention represented by the `tibble` class:

- each row is an observation;
- each column is a variable;
- each cell holds a single value.


### Structural Information

The most basic facts about a dataset are:

- the number of observations,
- the number and names of the variables,
- the structure (variable types).

```{r systemView, exercise = TRUE}
wvs7
```

```{r systemView-solution}
nrow(wvs7)   # number of rows
ncol(wvs7)   # number of columns
names(wvs7)  # variable (column) names
str(wvs7)    # variable names, types, and dimensions in one call
```


### Extracting Variables

Discussing the characteristics of a variable presupposes a way of referring to it. R, like other OOP systems, is well suited to working with multiple datasets and variables simultaneously — many datasets may be loaded at once, each stored in its own object.

OOP &rarr; multi-dataset analysis &rarr; dataset information + variable information

```{r variable, exercise = TRUE}
wvs7[, "country"]
```

```{r variable-solution}
wvs7$country
```


### Variable Characteristics

Variable extraction returns an ordinary vector, so all the vector techniques from the previous chapter apply. Two functions are particularly handy at this stage: `table()` for the empirical distribution of a categorical variable, and `summary()` for the standard numerical summary. R also supports specific summary statistics — sum, mean, median, min, max, variance, IQR, and so on — which we shall examine more closely in the next chapter.

```{r varCal, exercise = TRUE}
table(wvs7$age)
summary(wvs7$age)
```

For non-numeric variables, frequency tables remain the most informative summary:

```{r varCat}
table(wvs7$female)
table(wvs7$marital)
```

For factor-based variables, R additionally exposes the level structure:

```{r varLevel}
levels(wvs7$religious)
levels(wvs7$marital)
```


### Variable Attributes

Different variable classes carry different traits — a categorical variable has no meaningful mean, for instance — but every variable possesses certain general attributes such as length, class, type, and unique-value structure. The corresponding functions are themselves polymorphic.

```{r varAttr, exercise = TRUE}

```

```{r varAttr-solution}
length(wvs7$age)    # number of observations on age
unique(wvs7$age)    # the distinct values of age

summary(wvs7$age)   # five-number summary plus mean
class(wvs7$age)     # storage class
typeof(wvs7$age)    # underlying data type
```


### Variable Overview

```{r summary}
summary(wvs7$age)
summary(wvs7)
```


## Data Wrangling

Where data exploration looks at variables *through* a dataset, data wrangling rearranges the dataset *along* its variables. From a practical standpoint, we introduce wrangling through the **tidyverse** suite. Most of the operations below could be expressed in base R as well, but the tidyverse verbs read more naturally and chain together cleanly.


###

A brief note on the **tidyverse**:

- It is itself an R package.
- It is also a *family* of [R packages](https://www.tidyverse.org/packages/).
- Like the Marvel and DC universes, all tidyverse members operate within the same data structure, communicate with one another, and can be used together.


### Installation

```{r loadTidy, eval = FALSE}
install.packages("tidyverse")
library("tidyverse")
```


### The `dplyr` Package

`dplyr` is the tidyverse component devoted to data wrangling, faithful to the principle "one function does one thing".


### The Pipe

Before introducing the dplyr verbs, a word about the **pipe**. The native R pipe is `|>`; once `dplyr` is loaded, the more flexible magrittr pipe `%>%` also becomes available. The examples below use the latter.

A pipe links a sequence of operations on the same object — much like chaining moves in a fighting game.

###

Equally importantly, pipes make code easier to read: each step occupies its own line, in the order it is performed. For example, the dumpling-cooking process can be expressed without pipes as

```{r embeddedCode, eval = FALSE}
eat_dumpling <-
eat(
  dip(
    cook(
      fill(
        mix(
          meat,
          with = c(salt, soy_sauce, green_onion, ginger)
          ),
        with = wrapper
        ),
      in_water = boiled_water
      ),
    in_sauce = vinegar)
  )
```

###

With pipes:

```{r pipeCode, eval = FALSE}
eat_dumpling <-
mix(meat, with = c(salt, soy_sauce, green_onion, ginger)) %>%
  fill(with = wrapper) %>%
  cook(in_water = boiled_water) %>%
  dip(in_sauce = vinegar) %>%
  eat()
```

Keyboard shortcut for `%>%`:

* `Ctrl + Shift + M` (Windows / Linux)
* `Cmd + Shift + M` (macOS)


### Selecting Variables

`select(<data>, <var1>, <var2>, ...)`

The dataset has 24 variables, and some of the more interesting ones sit at the back where they are easy to overlook. Suppose we want a data frame containing only `country`, `age`, `education`, and `confidence_gov`:

```{r select, exercise = TRUE}
select(wvs7, country, age, education, confidence_gov)

# What if we want every variable about confidence, without listing each one explicitly?
```

```{r select-solution}
select(wvs7, country, age, education, starts_with("confidence"))
```

`starts_with()` has cousins `ends_with()` and `matches()`.

<div class="alert alert-info">

**Note**

Watch the third-person singular: `starts_with`, not `start_with`. 😝

</div>

###

To drop variables instead of keeping them, prepend `-`:

```{r deselect}
select(wvs7, -(country:education))
```

###

A close cousin of `select()` is `rename()`, which renames variables using the form `new_name = old_name`:

```{r rename, exercise = TRUE}
rename(wvs7, nationality = country)
```


### Sorting

`arrange(<data>, ...)`

Suppose we are curious about the youngest respondents' confidence in government, alongside their education and income.

```{r arrange, exercise = TRUE}
select(wvs7, age, confidence_gov, education, incomeLevel)
```

```{r arrange-solution}
select(wvs7, age, confidence_gov, education, incomeLevel) %>%
  arrange(age)
```

###

What about the oldest?

```{r arrangeDesc, exercise = TRUE}
select(wvs7, age, confidence_gov, education, incomeLevel) %>%
  arrange(desc(age))

## Among the youngest respondents, who has the most education?
```

```{r arrangeDesc-solution}
select(wvs7, age, confidence_gov, education, incomeLevel) %>%
  arrange(age, desc(education))
```


### Filtering by Value

While `select()` filters columns, `filter()` filters *rows* according to a logical condition. Continuing the previous example: among **American** respondents, who are the youngest, and what are their confidence in government, education, and income?

```{r filter, exercise = TRUE}
select(wvs7, age, confidence_gov, education, incomeLevel) %>%
  arrange(age)
```

```{r filter-solution}
select(wvs7, age, confidence_gov, education, incomeLevel, country) %>%
  filter(country == "United States") %>%
  arrange(age)

select(wvs7, age, confidence_gov, education, incomeLevel, country) %>%
  filter(country == "United States") %>%
  filter(age == min(age, na.rm = TRUE))
```


### Modifying and Creating Variables

In data analysis, we often need to transform or recompute variables; `mutate()` does the job. The English verb *to mutate* hints at the spirit of the function: not creation *ex nihilo* but transformation.

###

If we are interested in how income varies relative to education, we can construct a ratio:

```{r mutate, exercise = TRUE}
mutate(wvs7, ratio_incomeEdu = incomeLevel / (education + 1)) %>%
  select(country, incomeLevel, education, ratio_incomeEdu) %>%
  arrange(desc(ratio_incomeEdu))

# How would you express ratio_incomeEdu as a percentage?
```

```{r mutate-solution}
mutate(wvs7,
       ratio_incomeEdu = incomeLevel / (education + 1),
       ratio_incomeEdu = as.numeric(ratio_incomeEdu) %>%
         scales::percent()) %>%
  select(country, ratio_incomeEdu)
```


### Counting

`count()` produces frequency tables. We can use it to count the men and women in the sample.^[This kind of marginal table is a staple of census-style summaries.]

```{r count, exercise = TRUE}
wvs7 %>%
  count(female)

# How would you count men and women separately within each age group?
```

```{r count-solution}
wvs7 %>%
  count(age, female)
```


### Summarising

`summarise()` collapses many rows into a single summary row. To obtain the sample mean of age and education:

```{r summarise, exercise = TRUE}
wvs7 %>%
  summarise(age = mean(age, na.rm = TRUE),
            edu = mean(education, na.rm = TRUE))
```


### Grouped Operations

`group_by()` makes group-wise operations possible:

```{r ex_summaryG, exercise = TRUE}
wvs7 %>%
  group_by(female) %>%
  summarise(age = mean(age, na.rm = TRUE),
            edu = mean(education, na.rm = TRUE))
```

`group_by()` essentially attaches a group index to the dataset; subsequent operations are then performed within groups. The inverse operation is `ungroup()`.


## Bonus: `coalesce()`

How can we fill in the missing values of `x`, then merge `y` and `z` into a single variable?

```{r coalesce, exercise = TRUE}
df_toy <- data.frame(x = sample(c(1:2, NA, NA, NA)),
                     y = c(1, 2, NA, NA, 5),
                     z = c(NA, NA, 3, 4, 5))

df_toy
```

```{r coalesce-solution}
df_toy %>%
  mutate(x  = coalesce(x, 0L),
         yz = coalesce(y, z)) # Ta-da~~~
```


## Principles of Data Wrangling

Notice a feature shared by every operation above. What is it?

```{r rawData}
head(wvs7)
```

- <span style="color:red">Do not touch</span> the original data.
- Overwrite an object only with care.

```{r overwrite, exercise = TRUE}
wvs7 <- mutate(wvs7, female = as.numeric(female))
```


## Merging Datasets

### Direct Merging

The conditions for direct merging mirror those for matrix arithmetic — only commensurate shapes can be combined.

- Row-binding: the number of columns must <span style="color:red">match</span> (loosely speaking).
- Column-binding: the number of rows must <span style="color:red">match</span>.

A demonstration of each:

```{r bindRow, exercise = TRUE}
wvs7_us     <- filter(wvs7, country == "United States")
wvs7_russia <- filter(wvs7, country == "Russia")

# Build a US–Russia dataset

bind_rows(wvs7_us, wvs7_russia)

# What happens if the column counts do not match?
```

```{r bindRow-hint}
# Try this
bind_rows(tibble(x = 1:3), tibble(y = 1:4))
```

```{r bindCol, exercise = TRUE}
wvs7_conf  <- select(wvs7, starts_with("confidence"))
wvs7_trust <- select(wvs7, starts_with("trust"))

# Build a confidence-trust dataset

bind_cols(wvs7_conf, wvs7_trust)
```


### Index Merging (Joins)

Index merging refers to combining datasets on a shared index sequence (which may be any variable).

Let us first build two demonstration datasets:

1. an individual-level inequality-perception dataset;
2. a country-level demographic dataset.

If `wvs7_eq` is the survey data and `wvs7_country` the demographic data, our analysis requires merging the two to study the relationship between individual- and country-level variables.

```{r join, exercise = TRUE}
wvs7_eq <- select(wvs7, country, starts_with("equal")) %>%
  filter(country %in% unique(country)[1:2])

wvs7_country <- group_by(wvs7, country) %>%
  summarise(across(female:education, ~ mean(.x, na.rm = TRUE))) %>%
  ungroup() %>%
  filter(country %in% unique(country)[2:3])
```

```{r join-solution}
inner_join(wvs7_eq, wvs7_country)
left_join(wvs7_eq, wvs7_country)
right_join(wvs7_eq, wvs7_country)
full_join(wvs7_eq, wvs7_country)
```


## Summary

1. <span style="color:red">Think</span> before you act.
2. Combine the `dplyr` verbs flexibly and carefully.
    + Exploration: `head`, `tail`
        - Structure: `nrow`, `ncol`, `names`, `str`
        - Characteristics: `table`, `levels`
        - Attributes: `length`, `unique`, `summary`, `class`, `typeof`
    + Wrangling: the `dplyr` verbs
3. Principle of data wrangling: do not touch the original data.
4. Data merging
    - Direct: `bind_*`
    - Index-based: `*_join`
