---
title: "Data Wrangling"
subtitle: "Learn R with Dr. Hu and His Friends"
author: "Yue Hu (Department of Political Science, Tsinghua University)"
output:
  learnr::tutorial:
    language: "en"
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      out.width = "100%")

if (!require(pacman)) install.packages("pacman")

pacman::p_load(drhur, tidyverse)

data("wvs7", package = "drhur")
```


## Key Points

- Single-dataset operations: explore, select, filter, arrange, mutate
- Cross-dataset operations: direct merging, index merging

> **Research question:** Drawing on the WVS7 sample, which factors shape people's views of equality?


## Data Exploration

### A Glimpse of the Raw Data

```{r glimpse, exercise = TRUE, exercise.eval = TRUE}
wvs7
```

`wvs7` is stored as a **tibble** — a tidy data frame where each row is an observation, each column a variable, and each cell a single value.


### Structural Information

```{r structure}
nrow(wvs7)         # number of rows
ncol(wvs7)         # number of columns
head(names(wvs7))  # first six variable names
glimpse(wvs7)      # variable types and previews
```


### Extracting Variables

Two ways to access a variable: matrix-style brackets, or the `$` operator.

```{r variable}
# Bracket form (showing only the first 6 rows)
head(wvs7[, "country"])

# Dollar form (showing only first 10 unique values)
unique(wvs7$country) |> head(10)
```


### Variable Characteristics

For numeric variables, common summaries — mean, median, min, max, variance, IQR — are available directly.

```{r var-numeric}
mean(wvs7$age, na.rm = TRUE)
summary(wvs7$incomeLevel)
```

For categorical variables, `table()` returns the frequency distribution.

```{r var-categorical}
table(wvs7$age)
table(wvs7$female)
table(wvs7$marital)
```


## Data Wrangling with `dplyr`

The **tidyverse** is a family of R packages sharing a common philosophy and data structure. Its core components — **dplyr**, **ggplot2**, **tidyr**, **tibble** — work together seamlessly.


### The Pipe

R's native pipe `|>` connects sequential operations on the same object. Each step occupies its own line, in the order it is performed.

```{r pipe-nested, eval = FALSE}
# Traditional nested style: read inside-out, hard to follow
eatDumpling <-
eat(
  dip(
    cook(
      fill(
        mix(
          meat,
          with = c(salt, soy_sauce, green_onion, ginger)
          ),
        with = wrapper
        ),
      in_water = boiled_water
      ),
    in_sauce = vinegar)
  )
```

```{r pipe-style, eval = FALSE}
# Pipe style: read top-down, logic is clear
eatDumpling_pipe <-
mix(meat, with = c(salt, soy_sauce, green_onion, ginger)) |>
  fill(with = wrapper) |>
  cook(in_water = boiled_water) |>
  dip(in_sauce = vinegar) |>
  eat()
```

Keyboard shortcut for `|>` (or `%>%`): `Ctrl + Shift + M` (Windows / Linux), `Cmd + Shift + M` (macOS).


### Selecting Variables

`select()` keeps the listed columns:

```{r select, exercise = TRUE}
library(tidyverse)

# Pick five variables of interest
select(wvs7, equalIncentive, incomeLevel, confidence_gov, education, country)
```

Helpers like `starts_with()`, `ends_with()`, and `matches()` operate on column-name patterns. For example, the three corruption-perception variables all start with `corruption`:

```{r select-starts_with, exercise = TRUE}
# Pick all variables starting with "corruption"
select(wvs7, starts_with("corruption"))
```

To drop variables instead of keeping them, prepend `-`. We can combine `starts_with()` with `-` to drop a subset:

```{r select-deselect, exercise = TRUE}
# Keep corruption_local and corruption_civil, drop corruption_state
select(wvs7, starts_with("corruption"), -corruption_state)
```


### Renaming Variables

`rename()` uses the form `new_name = old_name`:

```{r rename, exercise = TRUE}
rename(wvs7, nationality = country)
```


### Filtering by Value

While `select()` keeps columns, `filter()` keeps rows that satisfy a logical condition.

```{r filter, exercise = TRUE}
# Respondents from Japan with high confidence in government
wvs7 |>
  filter(country == "Japan", confidence_gov >= 3)
```


### Arranging Rows

`arrange()` sorts by the listed variables (ascending by default; use `desc()` for descending).

```{r arrange, exercise = TRUE}
wvs7 |>
  select(age, confidence_gov, education, incomeLevel) |>
  arrange(age)
```


### Modifying and Creating Variables

`mutate()` modifies existing variables or creates new ones.

```{r mutate, exercise = TRUE}
# Average of two confidence variables
wvs7 |>
  mutate(avg_confidence = (confidence_gov + confidence_court) / 2) |>
  select(country, confidence_gov, confidence_court, avg_confidence)
```


## Merging Datasets

### Direct Merging

- **`bind_rows()`** stacks two data frames vertically. Mismatched columns are filled with `NA`.
- **`bind_cols()`** stacks two data frames side by side. The two must share row count.

```{r bind-rows, exercise = TRUE}
wvs7_jp <- filter(wvs7, country == "Japan")
wvs7_kr <- filter(wvs7, country == "South Korea")

bind_rows(wvs7_jp, wvs7_kr)
```

```{r bind-cols, exercise = TRUE}
wvs7_conf  <- select(wvs7, starts_with("confidence"))
wvs7_trust <- select(wvs7, starts_with("trust"))

bind_cols(wvs7_conf, wvs7_trust)
```


### Index Merging (Joins)

Index merging combines two datasets on a shared key.

<div class="alert alert-warning">

**Structural consistency**

Index merging requires that the index variables follow the same coding scheme. The variables need not contain identical sets of values, but their meanings must align.

</div>

We construct two demonstration datasets:

1. `wvs7_eq` — individual-level data on equality perceptions, restricted to Japan and South Korea.
2. `country_info` — country-level summaries of demographic variables, restricted to Japan and Thailand.

```{r join-prep, exercise = TRUE}
# Individual-level data: Japan and South Korea
wvs7_eq <- select(wvs7, country, equalIncentive, female, education) |>
  filter(country %in% c("Japan", "South Korea"))

# Country-level data: Japan and Thailand
country_info <- filter(wvs7, country %in% c("Japan", "Thailand")) |>
  group_by(country) |>
  summarise(across(female:education, \(x) mean(x, na.rm = TRUE))) |>
  ungroup() |>
  rename(nation = country)

wvs7_eq
country_info
```

dplyr provides four join types:

- **`left_join()`** — keep all rows of the left dataset; unmatched rows on the right become `NA`.
- **`right_join()`** — mirror image: keep all rows on the right.
- **`inner_join()`** — keep only rows whose key appears in both.
- **`full_join()`** — keep all rows from either side.

```{r join, exercise = TRUE}
# Run all four join types
wvs7_lj <- left_join(wvs7_eq,  country_info, by = join_by(country == nation))
wvs7_rj <- right_join(wvs7_eq, country_info, by = join_by(country == nation))
wvs7_ij <- inner_join(wvs7_eq, country_info, by = join_by(country == nation))
wvs7_fj <- full_join(wvs7_eq,  country_info, by = join_by(country == nation))

# Compare which countries each retained
wvs7_lj$country |> unique()
wvs7_rj$country |> unique()
wvs7_ij$country |> unique()
wvs7_fj$country |> unique()
```


## Section Exercise

Build a single dplyr pipeline on `wvs7` that:

1. Creates a new variable `avg_confidence` — the mean of `confidence_gov` and `confidence_court`.
2. Keeps respondents from Japan, aged strictly between 40 and 60, with `avg_confidence` above 3.
3. Sorts the result by `avg_confidence` in descending order.
4. Returns only `age`, `incomeLevel`, `equalIncentive`, and `avg_confidence`.

```{r final_exercise, exercise = TRUE}

```

```{r final_exercise-solution}
wvs7 |>
  mutate(avg_confidence = (confidence_gov + confidence_court) / 2) |>
  filter(
    country == "Japan",
    age > 40,
    age < 60,
    avg_confidence > 3
  ) |>
  arrange(desc(avg_confidence)) |>
  select(age, incomeLevel, equalIncentive, avg_confidence)
```


## Summary

- **Data exploration**: `nrow()`, `ncol()`, `glimpse()`, `names()`, `unique()`, `table()`
- **Data wrangling**
    - Pipe: `|>`
    - Select columns: `select()`
    - Rename: `rename()`
    - Filter rows: `filter()`
    - Arrange: `arrange()`
    - Modify or add: `mutate()`
- **Data merging**
    - Row-binding: `bind_rows()`
    - Column-binding: `bind_cols()`
    - Joins: `left_join()`, `right_join()`, `inner_join()`, `full_join()`
