---
title: "Programming Fundamentals and Data I/O"
subtitle: "Learn R with Dr. Hu and His Friends"
author: "Yue Hu (Department of Political Science, Tsinghua University)"
output:
  learnr::tutorial:
    language: "en"
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
chooseCRANmirror(graphics = FALSE, ind = 1)
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      out.width = "100%")

if (!require(pacman)) install.packages("pacman")

pacman::p_load(
  lubridate,
  here,
  rio,
  tidyverse,
  drhur
)
```


## Key Points

- Foundational concepts (object-oriented programming)
- Basic syntax (functions, packages, assignment, naming)
- Data types
- Data attributes
- Data input and output


## Object-Oriented Programming

The previous chapter introduced how R realises the content of mathematics and statistics textbooks. The present chapter shifts perspective and presents R's foundational concepts from the standpoint of programming. On that basis, we take the first practical step in working with R: importing and exporting data ŌĆö the two ends of any data analysis pipeline. We shall also examine how to interrogate basic features of a dataset, such as its size, its variable count, and its number of observations.

We begin by sketching R's programming character.

###

**Object-Oriented Programming** (OOP) is a widely adopted architecture in modern computer programming. Compared with traditional procedure-oriented programming, OOP organises and encapsulates data so that more complex systems and demands can be handled cleanly. Many widely known languages ŌĆö C++, Java, Ruby, Python ŌĆö adopt OOP. R does the same, although in a rather idiosyncratic form (more on this when we discuss functions). Understanding OOP makes much of R's behaviour easier to reason about.

OOP rests on two basic notions: **classes** and **objects**.

- **Class:** a user-definable structure of an *object*, together with its *properties* and *methods*.
- **Object:** a concrete instance of a particular class ŌĆö much as "China" is to "country", or "R" is to "letter". Each instance carries the properties defined for its class.
- **Method:** a command defined within a class that expresses some characteristic action of objects of that class. If a person eats, drinks, sleeps, and so on, then "eating, drinking, sleeping" are the person's methods.
- **Property:** an attribute used to describe some aspect of an object ŌĆö hair colour, skin colour, height, weight, and so forth, for a person. All objects of a given class share the same set of properties, though their values may differ. Think of properties as the parameters of an object.

###

Beyond these, OOP languages typically share four additional traits.

- **Encapsulation.** All properties and methods of an object are bundled together. Other objects cannot directly inspect or alter the inner state of an object ŌĆö they may only influence it. One cannot turn an ectotherm into an endotherm, but one can affect a lizard's body temperature by changing the ambient temperature of its tank.
- **Polymorphism.** A single command can act differently on different object types. The most familiar example in R is `summary()`, which adapts its output to whatever class of object it receives.

```{r polymorphism, exercise = TRUE}
summary(wvs7)
summary(wvs7$age)
```

### Polymorphism in action

R's polymorphism allows generic functions to handle object types they have not been explicitly programmed for. Beyond `summary()`, `plot()` is another classic polymorphic function. Apply `plot()` to two different kinds of object and observe how the output differs.

```{r polymorphism_exercise, exercise = TRUE}
plot(wvs7$age)
plot(wvs7$age, wvs7$incomeLevel)
```

###

- **Inheritance.** All subclasses of a parent class automatically possess the parent's traits. By analogy: if humans have two legs, every individual human inherits that trait.
- **Safety.** When a command is applied to an object, R checks the object's class. If the class lies outside the command's defined scope, R halts and returns an error.


### Functions: Imperative Sentences with the Adverbial Postposed

R is built from two kinds of building blocks: functions and objects. A **function** can be thought of as an imperative sentence in which the adverbial phrase appears *after* the verb ŌĆö exactly opposite to typical English word order. The general form is

> `<function_name>(<target_data>, <condition_1>, <condition_2>, ...)`

The OOP creed expressed by this design is straightforward: *do not touch the data; let the function act on it.*

```{r funIllustration, eval = FALSE}
light <- function(finger){
  shadow <- finger + 5
  shadow
}
handShadow <- light(finger = 3)
handShadow
```


### Packages

A **package** is a curated collection of functions ŌĆö much like an app on a smartphone. A working snapshot of CRAN holds tens of thousands of registered packages (and many more on GitHub). Two common ways to install packages are:

```{r install, eval = FALSE}
install.packages("drhur")                          # from CRAN
devtools::install_github("sammo3182/drhur")         # from GitHub
```


### The Assignment Arrow `<-`

The arrow `<-` is the assignment operator and is shorthand for the `assign()` function's default behaviour. The general form is

> `<name> <- <object>`

```{r object}
aValidObject <- 1:5
aValidObject
```


### `->`, `=`, `<<-`

R supports four ways to assign:

1. `assign()`
2. `<-`
3. `<<-`
4. `=`


### Why `<-`

`<-` is preferred for several reasons:

- *Direction is intuitive.* The arrow makes the data flow visually obvious ŌĆö from the right-hand object into the left-hand name.

```{r arrow}
a <- 12
25 -> b   # the rightward arrow also works
```

- *No confusion with equality.* Using `<-` for assignment leaves `=` and `==` unambiguous.
- *A keyboard shortcut exists.*
    - PC: `Alt + -`
    - Mac: `Option + -`


### When to Use Each

- `=` ŌĆö when you do *not* want to create a permanent object, for instance when supplying named arguments inside a function call.

```{r sideEffect, exercise = TRUE}
median(y <- 1:10); y
median(x = 1:10); x
```

- `<<-` ŌĆö to refer to a variable in a parent scope, most often inside closures.

```{r superassign}
new_counter <- function() {
  i <- 0
  function() {
    # do something useful, then ...
    i <<- i + 1
    i
  }
}
```


### Naming Rules

Object names in R must satisfy several constraints:

1. They must not begin with a digit (incorrect: `1stday`).
2. Apart from `.` and `_`, no special characters are allowed (incorrect: `M&M`).
3. Names are case-sensitive (`X != x`); recall that `!` denotes "not" while `!=` denotes "not equal".
4. Avoid overwriting built-in names (avoid `list <- c(1:5)`).
5. Choose names that convey meaning.

Try creating one valid and one invalid object below.

```{r objectEg, exercise = TRUE}

# An invalid object name (begins with a digit)
# 5var_name <- data_frame(wvs7$education)


# A valid object name
var_name5 <- data_frame(wvs7$education)

```


### Practice

```{r vname, echo = FALSE}
question("Which of the following are valid variable names?",
  answer("my_data_frame <- data_frame(wvs7$education)", correct = TRUE),
  answer("mydata&frame <- data_frame(wvs7$education)"),
  answer("MyDataFrame <- data_frame(wvs7$education)", correct = TRUE),
  answer("1data_frame <- data_frame(wvs7$education)"),
  incorrect = "Not quite ŌĆö recall that names cannot begin with a digit and may not contain special characters other than `.` and `_`.")
```


## Data Input

### Built-in Datasets

R ships with a substantial collection of demonstration datasets, listed by `data()`. Many R packages also include their own datasets; once a package is loaded, its datasets become available in the same way.

```{r mtcars, exercise = TRUE}
data()
```

###

Pick a dataset listed by `data()`, load it, and use `summary()` to inspect its variables.

```{r mtcars_exercise, exercise = TRUE}


```

```{r mtcars_exercise-solution}
# An example
data(uspop)
summary(uspop)
```


### Directly Readable Formats

Several common file formats can be read with base R alone:

- `.RDS` ŌĆö single R object
- `.RData` ŌĆö multiple R objects
- `.txt` ŌĆö plain-text tabular data
- `.csv` ŌĆö comma-separated values

> Syntax: `<name> <- <reader>(<path>)`

```{r input, eval = FALSE}
df_rds <- readRDS("aDataset.rds")
df_txt <- read.table("D:/aDataset.txt")
df_csv <- read.csv("./aDataset.csv")
```


### Reading via Packages

For other file formats, specialised packages offer dedicated readers. Load them with `library` or `require` and then call the appropriate function.

```{r read_pkgs, eval = FALSE}
# SPSS, Stata, SAS
library(haven)
df_spss  <- read_spss("<FileName>.sav")
df_stata <- read_dta("<FileName>.dta")
df_sas   <- read_sas("<FileName>.sas7bdat")

# Tabular files (fast import)
library(readr)
df_csv   <- read_csv("<FileName>.csv")
df_table <- read_table("<FileName>.txt")

# Excel
library(readxl)
df_excel  <- read_excel("<FileName>.xls")
df_excel2 <- read_excel("<FileName>.xlsx")

# JSON
library(rjson)
df_json <- fromJSON(file = "<FileName>.json")

# XML / HTML
library(xml2)
df_xml  <- read_xml("<url>")
df_html <- rvest::html_table(rvest::read_html("<url>"))
```

###

The Swiss-army knife of data import is **rio**, whose `import()` function dispatches to the appropriate reader based on the file extension:

```{r rio, eval = FALSE}
library(rio)
df_anything <- import("<AnyTypeOfData>")
```


## Data Types

1. Vector
2. Matrix
3. Data frame
4. List
5. Array


### Vectors

The function `c()` ("combine") is the workhorse for constructing vectors. A vector's elements must share the same data type ŌĆö numeric, character, logical, factor, or otherwise.

#### Numeric Vectors

```{r numeric, exercise = TRUE}
vec_integer <- c(1, -2, NA)
vec_double  <- c(1.5, -2.34, 1/3)
```

Two notes:

1. `NA` stands for *not available* ŌĆö R's representation of a missing value.
2. All elements of a single vector must share one type; if you mix types, R will silently coerce them to the most general one.

###

Generate a vector containing every even integer from 1 to 100.

```{r numeric_exercise, exercise = TRUE}
# hint: help(seq)

```

```{r numeric_exercise-solution}
seq(2, 100, by = 2)
```


#### Character Vectors

```{r character, exercise = TRUE}
vec_chr <- c("R", "is", "powerful ŌĆö and so are you.")
```

###

Generate the sequence of lowercase Roman letters from `a` to `z`.

```{r character_exercise, exercise = TRUE}
vec_letters <- c("a", "b", "c", "d", "e")
```

```{r character_exercise-solution}
letters[1:26]
```


#### Logical Vectors

```{r logic, exercise = TRUE}
vec_tf <- c(TRUE, TRUE, FALSE)
vec_tf
# c(TRUE, TRUE, FALSE) == c(1, 1, 0)
```

###

Suppose `x` is a vector containing `(1, 1, 0)`. Convert it into a logical vector.

```{r logic_exercise, exercise = TRUE}

```

```{r logic_exercise-solution}
x <- c(1, 1, 0)
x <- as.logical(x)
```


#### Factor Vectors

A **factor** stores a categorical variable with a fixed set of allowed values (its *levels*). An *ordered* factor additionally records a meaningful order among the levels.

```{r factor, exercise = TRUE}
vec_fac <- factor(c(1, 2, 2, 3))
```

```{r factor-hint}
vec_ord  <- ordered(c(1, 2, 2, 3))
vec_fac2 <- factor(c(1, 2, 2, 3),
                   levels = c(3, 2, 1),
                   labels = c("Apple", "Pear", "Orange"))
```

###

When a new dataset arrives, the first task is to develop a rough sense of what is inside. Inspect the variables in `wvs7`:

```{r factor_exercise1, exercise = TRUE}



```

```{r factor_exercise1-solution}
str(wvs7)
```

###

Examine the class of `wvs7$incomeLevel`:

```{r factor_exercise2, exercise = TRUE}



```

```{r factor_exercise2-solution}
class(wvs7$incomeLevel)
```

###

List the values of `incomeLevel` and the frequency of each:

```{r factor_exercise3, exercise = TRUE}



```

```{r factor_exercise3-solution}
table(wvs7$incomeLevel)
```


#### Date and Time Vectors

R offers two representations of time:

- `as.POSIXct()` ŌĆö numeric input, integer storage; counts the seconds since the UNIX epoch (1970-01-01 00:00:00).
- `as.POSIXlt()` ŌĆö character input, list storage; each component (year, month, day, and so on) is an element of the list.

```{r time_POSIXct_POSIXlt}
# The difference between as.POSIXct and as.POSIXlt
ct <- as.POSIXct("2023-03-20 10:11:12")
lt <- as.POSIXlt("2023-03-20 10:11:12")
```

```{r time_POSIXctandPOSIXlt}
unlist(ct)
unlist(lt)
```

```{r time, exercise = TRUE}
Sys.time()    # current date and time (with time zone)
today()       # today's date (year, month, day)
now()         # today's date and time
# CST is the time zone of the machine running this code

# The full pack
time1 <- Sys.time()
time2 <- as.POSIXlt(Sys.time())
time2$wday    # day of the week

## What if you only care about the date?
```

```{r time-hint}
Sys.Date()
date1 <- as.Date("2019-01-02")
class(date1)
```

###

The Swiss-army knife for date handling is **lubridate**:

```{r date, exercise = TRUE}
library(lubridate)

ymd("20221016")
mdy("10-16-2022")
dmy("16/10/2022")
ymd_hms("2022-10-16 09:00:00", tz = "Etc/GMT+8")
```

```{r date-hint}
OlsonNames()
```

###

When date strings appear in inconsistent orders, such as

```
x <- c("20190101", "01012019", "021901")
```

how would you parse them?

```{r date_exercise, exercise = TRUE}
# help(parse_date_time)

```

```{r date_exercise-solution}
x <- c("20190101", "01012019", "021901")
parse_date_time(x, orders = c("ymd", "dmy", "dym"))
```


### Matrices

For matrices, see the previous chapter ŌĆö `drhur("calculate")`.


### Arrays

An **array** is, as the name suggests, an "array of columns" that can store data of more than two dimensions. Use `array()` to create one.

```{r array}
# Create two vectors of different lengths
vector1 <- c(5, 9, 3)
vector2 <- c(10, 11, 12, 13, 14, 15)

# Feed them into an array
result <- array(c(vector1, vector2), dim = c(3, 3, 2))
result
```


### Lists

A **list** is a heterogeneous "string" of objects ŌĆö its elements may belong to different types or even differ in length:

```{r list}
ls_monks <- list(name   = c("Wukong Sun", "Sanzang Tang", "Wuneng Zhu", "Wujing Sha"),
                 power  = c(100, 20, 90, 40),
                 buddha = c(TRUE, TRUE, FALSE, FALSE))

ls_monks
```


### Data Frames

A **data frame** is a special kind of list (or, equivalently, of matrix) in which:

- columns are *variables*, each of equal length;
- rows are *observations*.

In R:

```{r dataframe}
df_toy <- data.frame(female = c(0, 1, 1, 0),
                     age    = c(29, 39, 38, 12),
                     name   = c("Iron Man", "Black Widow",
                                "Captain Marvel", "Captain America"))

df_toy
```


## Data Attributes

1. `class`, `typeof` ŌĆö query the variable's storage class and underlying type.
2. `nchar` ŌĆö get the length of a character string.
3. `levels` ŌĆö get or set the levels of a factor.
4. `nrow` ŌĆö return the number of rows of a matrix or data frame.
5. `ncol` ŌĆö return the number of columns of a matrix or data frame.
6. `dim` ŌĆö return the dimensions (the subspace spanned by the column vectors).

```{r attribute, exercise = TRUE}
vec_integer <- c(1, -2, NA)

vec_double <- c(1.5, -2.34, 1/3)

vec_chr <- c("R", "is", "powerful ŌĆö and so are you.")

vec_fac <- factor(c(1, 2, 2, 3))

ls_monks <- list(name   = c("Wukong Sun", "Sanzang Tang", "Wuneng Zhu", "Wujing Sha"),
                 power  = c(100, 20, 90, 40),
                 buddha = c(TRUE, TRUE, FALSE, FALSE))

df_toy <- data.frame(female = c(0, 1, 1, 0),
                     age    = c(29, 39, 38, 12),
                     name   = c("Iron Man", "Black Widow",
                                "Captain Marvel", "Captain America"))

class(vec_double)
typeof(vec_integer)

nchar(vec_chr)
levels(vec_fac)

length(vec_double)
length(ls_monks)
length(df_toy)

nrow(df_toy)
ncol(df_toy)
dim(df_toy)
```


### Practice

Convert the following vector into numeric form:

```
c(FALSE, TRUE)
```

```{r attribute_exercise1, exercise = TRUE}
# help(as.numeric)

```

```{r attribute_exercise1-solution}
as.numeric(c(FALSE, TRUE))
```

###

In `wvs7`, the variable `female` takes the logical values `TRUE` / `FALSE`.

For analysis, it is often more convenient to recode logical values as numeric 0 / 1.

```{r attribute_exercise2, exercise = TRUE}

```

```{r attribute_exercise2-solution}
as.numeric(wvs7$female) - 1
```


## Data Output

> Syntax: `<writer>(<object_to_save>, file = <path>)`

### Storing as Native R Data

```{r saving, eval = FALSE}
saveRDS(df_toy, file = "df_toy.rds")
save(df_toy, ls_monks, file = "test.RData")
```


### Storing as CSV

```{r savingCSV, eval = FALSE}
write.csv(df_toy, file = "toy.csv")
```

<div class="alert alert-info">

**A note on encoding**

If your dataset contains Chinese (or other non-Latin) text, `write.csv()` may produce mojibake (garbled characters) on some systems. Specifying `fileEncoding = "UTF-8"` or saving with `readr::write_csv()` typically resolves the issue.

</div>

###

You may export to Stata, SPSS, SAS, Excel, JSON, MATLAB, HTML, and many other formats ŌĆö but should you? The native R formats are usually faster and more compact:

STATA (.dta, < 14): 3.16 G = R (.rds): 0.05 G

| **Method**       | **Average Time** | **Minimum** | **Maximum** |
|:-----------------|:----------------:|:-----------:|:-----------:|
| `base::readRDS`    |       19.65    |    18.64    |    21.01    |
| `fst::read_fst`    |        1.39    |     0.56    |     3.41    |
| `haven::read_sav`  |      104.78    |   101.00    |   111.85    |
| `qs::qread`        |        3.33    |     3.00    |     4.24    |

: Average time (in seconds) to read GSS data in R.


| **Method**      | **Average Time** | **Minimum** | **Maximum** | **File Size** |
|:----------------|:----------------:|:-----------:|:-----------:|:-------------:|
| `base::saveRDS`   |       98.36    |    93.09    |   103.24    |    30.9 MB    |
| `fst::write_fst`  |        2.70    |     1.86    |     4.05    |   122.1 MB    |
| `qs::qsave`       |        5.03    |     4.35    |     6.62    |    44.6 MB    |

: Average time (and file size) to write GSS data in R.


## Summary

- **Foundational concepts**
    - OOP
    - Functions and packages
    - Assignment
- **Data input**
    - Built-in datasets and CSV
    - Package-mediated readers
- **Data types**
    - Vectors
    - Matrices
    - Arrays
    - Lists
    - Data frames
- **Data attributes**
- **Data output**
    - Native R formats
    - Other tabular formats
