---
title: "Programming Fundamentals and Data I/O"
subtitle: "Learn R with Dr. Hu and His Friends"
author: "Yue Hu (Department of Political Science, Tsinghua University)"
output:
  learnr::tutorial:
    language: "en"
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
chooseCRANmirror(graphics = FALSE, ind = 1)
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      out.width = "100%")

if (!require(pacman)) install.packages("pacman")

pacman::p_load(lubridate, tidyverse, drhur)
```


## Key Points

- Foundational concepts (OOP, classes, objects)
- Data types
- Data attributes
- Data input and output


## Object-Oriented Programming

**Object-Oriented Programming** (OOP) organises and encapsulates data through *classes* and *objects*. Many widely known languages — C++, Java, Python — adopt OOP, and R does the same.

- **Class:** a user-definable structure of an object, together with its properties and methods.
- **Object:** a concrete instance of a class — "China" is to "country", "R" is to "letter".
- **Method:** a command defined within a class.
- **Property:** an attribute of an object — like hair colour for a person.

OOP languages also typically share **encapsulation**, **polymorphism**, **inheritance**, and **safety**.

###

**Polymorphism** lets one command act differently on different object types. The most familiar example is `summary()`:

```{r polymorphism, exercise = TRUE}
summary(wvs7)
summary(wvs7$age)
```


### Functions

A **function** can be thought of as an imperative sentence with the adverbial postposed:

> `<function_name>(<target_data>, <condition_1>, <condition_2>, ...)`

The OOP creed: *do not touch the data; let the function act on it*.

```{r funIllustration, eval = FALSE}
light <- function(finger){
  shadow <- finger + 5
  shadow
}
handShadow <- light(finger = 3)
handShadow
```


### Packages

A **package** is a curated collection of functions. Two common installation methods:

```{r install, eval = FALSE}
install.packages("drhur")                          # from CRAN
devtools::install_github("sammo3182/drhur")         # from GitHub
```


### Exercise 1

Show how to install three packages — `regioncode`, `dotwhisker`, `drhur` — in a single call.

```{r install_exercise, exercise = TRUE}
?install.packages

```

```{r install_exercise-solution}
install.packages(c("regioncode", "dotwhisker", "drhur"))
```


### Exercise 2

`plot()` is another polymorphic function. Apply it to one variable, then to two.

```{r polymorphism_exercise, exercise = TRUE}

```

```{r polymorphism_exercise-solution}
plot(mtcars$mpg)              # one variable: index vs value
plot(mtcars$mpg, mtcars$wt)   # two variables: scatter plot
```


### The Assignment Arrow `<-`

`<-` is R's standard assignment operator, equivalent to `assign()`:

> `<name> <- <object>`

```{r object}
aValidObject <- 1:5
aValidObject
```

R supports four assignment symbols: `assign()`, `<-`, `<<-`, `=`. `<-` is preferred — direction is intuitive, and it does not collide with `=` or `==`. Keyboard shortcut: **Alt + -** (Windows/Linux), **Option + -** (macOS).

```{r arrow}
a <- 12
25 -> b   # the rightward arrow also works
```


### Naming Rules

Object names must satisfy:

1. Not begin with a digit (incorrect: `1stday`).
2. Contain no special characters except `.` and `_` (incorrect: `M&M`).
3. Are case-sensitive (`X != x`).
4. Avoid overwriting built-in names (avoid `list <- c(1:5)`).
5. Convey meaning.


### Exercise 3

Why does the following line fail?

```{r naming_exercise, exercise = TRUE}
5var_name <- 1:5
```

```{r naming_exercise-solution}
# Object names cannot begin with a digit. "5var_name" violates Rule 1.
# A valid alternative:
var_name5 <- 1:5
var_name5
```


## Records (Data Storage)

### Vectors

`c()` builds vectors. A vector's elements must share the same type.

#### Numeric Vectors

```{r numeric}
vec_integer <- c(1, -2, NA)
vec_double  <- c(1.5, -2.34, 1/3)
```

`NA` stands for *not available* — R's representation of a missing value.


#### Character Vectors

```{r character}
vec_chr <- c("R", "is", "useful.")
vec_chr
```


### Exercise 4

Generate the lowercase Roman alphabet `a` to `e` in two ways and check that they match.

```{r character_exercise, exercise = TRUE}

```

```{r character_exercise-solution}
vec_letters_raw   <- c("a", "b", "c", "d", "e")
vec_letters_quick <- letters[1:5]

# Check they're identical
identical(vec_letters_raw, vec_letters_quick)
```


#### Logical Vectors

```{r logic}
vec_tf <- c(TRUE, TRUE, FALSE)
vec_tf
```


#### Factor Vectors

A **factor** stores a categorical variable with a fixed set of allowed values (its *levels*).

```{r factor}
vec_fac  <- factor(c(1, 2, 2, 3))
vec_ord  <- ordered(c(1, 2, 2, 3))
vec_fac2 <- factor(c(1, 2, 2, 3),
                   levels = c(3, 2, 1),
                   labels = c("Apple", "Pear", "Orange"))

vec_fac
vec_ord
vec_fac2
```


#### Date and Time Vectors

R offers two representations of time:

- `as.POSIXct()` — numeric storage, seconds since 1970-01-01.
- `as.POSIXlt()` — list storage, each component as a separate element.

```{r time-classes}
ct <- as.POSIXct("2026-04-13 10:11:12")
lt <- as.POSIXlt("2026-04-13 10:11:12")

unlist(ct)
unlist(lt)
```

The Swiss-army knife for dates is **lubridate**:

```{r lubridate}
ymd("20221016")
mdy("10-16-2022")
dmy("16/10/2022")
ymd_hms("2022-10-16 09:00:00", tz = "Etc/GMT+8")
```


### Exercise 5

When date strings appear in inconsistent orders, parse them with `parse_date_time()`:

```{r date_exercise, exercise = TRUE}
raw_dates <- c("20190101", "01012019", "021901")

```

```{r date_exercise-solution}
library(lubridate)
raw_dates <- c("20190101", "01012019", "021901")

parsed_dates <- parse_date_time(raw_dates, orders = c("ymd", "dmy", "mdy"))
parsed_dates
```


### Matrices, Arrays, Lists, and Data Frames

**Matrices** were introduced in `drhur("calculate")`. **Arrays** generalise matrices to higher dimensions:

```{r array}
vector1 <- c(5, 9, 3)
vector2 <- c(10, 11, 12, 13, 14, 15)

result <- array(c(vector1, vector2), dim = c(3, 3, 2))
result
```

A **list** is a heterogeneous container; elements may differ in type or length:

```{r list}
ls_monks <- list(
  name   = c("Wukong Sun", "Sanzang Tang", "Wuneng Zhu", "Wujing Sha"),
  power  = c(100, 20, 90, 40),
  buddha = c(TRUE, TRUE, FALSE, FALSE)
)

ls_monks
```

A **data frame** is a special list (or matrix) where each column is a variable of equal length and each row is an observation:

```{r dataframe}
df_toy <- data.frame(
  female = c(0, 1, 1, 0),
  age    = c(29, 39, 38, 12),
  name   = c("Iron Man", "Black Widow", "Captain Marvel", "Captain America")
)

df_toy
```


### Data Attributes

Functions for inspecting variables: `class()`, `typeof()`, `length()`, `dim()`, `nrow()`, `ncol()`, `levels()`, `nchar()`, `summary()`, `head()`.

```{r attribute}
class(vec_double)
typeof(vec_integer)
length(vec_double)
dim(df_toy)
nrow(df_toy)
ncol(df_toy)
```


### Exercise 6

The `wvs7` dataset (a 2 % sample of WVS Wave 7, included in `drhur`) holds 1264 observations on 24 variables. Inspect its basic structure.

```{r wvs_exercise, exercise = TRUE}
data("wvs7", package = "drhur")

```

```{r wvs_exercise-solution}
data("wvs7", package = "drhur")

# Basic situation
class(wvs7)         # "tbl_df" "tbl" "data.frame"
dim(wvs7)           # 1264   24

# Variable attributes
head(wvs7$marital)  # first few values
levels(wvs7$marital)  # factor levels

# Structure verification
is.data.frame(wvs7)  # TRUE
is.list(wvs7)        # TRUE (a data frame is a list)
is.matrix(wvs7)      # FALSE
```


## Data Input and Output

### Built-in Datasets

R ships with demonstration datasets, listed by `data()`. Many R packages include their own.

```{r datasets, eval = FALSE}
data()
```


### Direct Reading

```{r input, eval = FALSE}
df_rds <- readRDS("aDataset.rds")
df_txt <- read.table("D:/aDataset.txt")
df_csv <- read.csv("./aDataset.csv")
```


### Reading via Packages

```{r read_pkgs, eval = FALSE}
library(haven)
df_spss  <- read_spss("<FileName>.sav")
df_stata <- read_dta("<FileName>.dta")

library(readxl)
df_excel <- read_excel("<FileName>.xlsx")

library(rio)
df_anything <- import("<AnyTypeOfData>")
```


### Storing Data

```{r save, eval = FALSE}
saveRDS(df_toy, file = "df_toy.rds")
save(df_toy, ls_monks, file = "test.RData")
write.csv(df_toy, file = "toy.csv")
```


### Exercise 7

Load the built-in `uspop` dataset and inspect its statistical summary.

```{r uspop_exercise, exercise = TRUE}

```

```{r uspop_exercise-solution}
data(uspop)
summary(uspop)
```


## Summary

- **Foundational concepts**: OOP (classes, objects); functions and packages; assignment.
- **Data types**: vectors (numeric, character, logical, factor, date/time); matrices, arrays, lists, data frames.
- **Data attributes**: `class()`, `dim()`, `summary()`, `head()`.
- **Data I/O**: `data()`, `read.csv()`, `readRDS()`; `saveRDS()`, `write.csv()`.
