---
title: "Basic Mathematical Concepts"
subtitle: "Learn R with Dr. Hu and His Friends"
author: "Yue Hu (Department of Political Science, Tsinghua University)"
output:
  learnr::tutorial:
    language: "en"
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      out.width = "100%")
```

## Key Points

- Basic algebra
- Vectors
- Matrices and equations


## Basic Operations

R's arithmetic symbols match those of mathematics: `+`, `-`, `*`, `/`.

```{r algebra-basic}
# The four basic operations
1 + 2
3 - 4
5 * 6
7 / 8

# Compound expressions
(2 + 4) / (5 - 3)
```

<div class="alert alert-info">

**A note on whitespace**

Spaces between numbers and operators are added for readability — R itself is indifferent to whitespace. The expression below runs perfectly:

```{r whitespace-demo}
(2+4) /  (  5-   3        )
```

For inherited or third-party code, packages such as **styler** and **air** can reformat code in a single click.

</div>

###

R also supports more complex arithmetic: powers (`^`), modulo (`%%`), logarithms (`log()`, `log10()`), exponentials (`exp()`), square root (`sqrt()`), and absolute value (`abs()`).

```{r complex-math}
10^2
5 %% 3      # remainder
log(11)
log10(100)
exp(12)

# Two ways to take a square root
9^(1/2)
sqrt(9)

# Absolute value
abs(-9)
```

###

Trigonometric functions:

```{r triangle}
sin(pi)
cos(pi)
tan(pi / 2)

acos(1)
asin(1)
atan(0.5)
```

###

R pre-defines several useful constants. `pi` is one. Others include `letters`, `LETTERS`, `month.name`, `month.abb`. Try them below:

```{r reserved, exercise = TRUE}

```

```{r reserved-solution}
letters
month.name
```


### Scientific Notation

`tan(pi / 2)` above returned `1.633124e+16`. This is **scientific notation**: a real coefficient $a$ ($1 \leq |a| < 10$) times an integer power of 10. The letter `e` connects the coefficient and the exponent. So $200 = 2 \times 10^2$ is written `2e2`, and $0.002 = 2 \times 10^{-3}$ is `2e-3`.


### Boolean Algebra

R uses `TRUE` / `FALSE` (or `T` / `F`) for the two truth values. In numeric contexts, `FALSE` becomes `0` and `TRUE` becomes `1`. The three Boolean connectives are `&` (and), `|` (or), `!` (not). The Boolean empty set is `NULL`.

Boolean expressions pair with comparison operators: `==`, `!=`, `>`, `<`, `>=`, `<=`.

```{r boolean}
# Equalities
1 == 2
1 == TRUE   # TRUE and FALSE behave as 1 and 0 with logical attributes
0 != F

# Inequalities
1 != 2
1 > 2
1 < 2
1 >= 2
1 <= 2

# Boolean operations
3 > 2 & 2 > 1
12 & 2 > 2
-3 | 2 > 2
```

In `3 > 2 & 2 > 1`, R first evaluates the comparisons, then the connective. When a single value enters a Boolean expression — as in `12 & 2 > 2` — R treats it as a self-equality check.

<div class="alert alert-info">

**Why the double equal sign?**

In R, the single `=` is reserved for assignment, so equality requires `==`. This keeps it consistent with `!=`, `>=`, `<=`. Forgetting and writing `=` is normal at first — practice will fix it.

</div>


## Vectors

In R, `c()` ("concatenate") builds vectors. For consecutive integers, the colon `:` is a shortcut. The sequence function `seq()` is the third option.

```{r vector}
# Method 1: c()
a  <- c(11, 12, 13, 14, 15)

# Method 2: colon (consecutive integers only)
a1 <- 11:15

# Method 3: seq()
a2 <- seq(from = 11, to = 15, by = 1)
a3 <- seq(11, 15)

# All three match the original
a == a1
a == a2
a == a3
```


### Indexing

Square brackets `[]` extract elements from a vector.

```{r indexing}
a <- c(11, 12, 13, 14, 15)

# Single element
a[2]
a[5]

# Multiple elements
a[c(3, 4)]
```

<div class="alert alert-info">

**The assignment arrow `<-`**

`<-` is R's standard assignment operator. The arrow points left: the right-hand value is *stored into* the left-hand name.

R is essentially the only popular language using `<-`; others use `=`. The reason traces back to APL (A Programming Language) in the 1960s, whose dedicated keyboard had a left-arrow key in the upper-right corner. APL faded, but the `<-` habit persisted into S, and then into R.

Keyboard shortcut: **Alt + -** on Windows / Linux, **Option + -** on macOS.

</div>


### Vector Arithmetic

Addition and subtraction combine corresponding elements:

$$\vec{a} \pm \vec{b} = (a_1 \pm b_1,\; a_2 \pm b_2,\; \ldots,\; a_n \pm b_n).$$

```{r vec-add}
a <- c(11, 12, 13, 14, 15)

# Vector ± scalar
b <- a + 4
c <- a - 0.5

# Vector ± vector
a + b
a - c
```

###

For multiplication, `*` is the **element-wise (Hadamard) product**, `%*%` is the **inner (dot) product**, and `%o%` (or `outer()`) is the **outer (tensor) product**.

```{r vec-mult}
a <- c(11, 12, 13, 14, 15)
b <- a + 4

# Element-wise (Hadamard) product
a * b

# Inner (dot) product
a %*% b

# Outer (tensor) product
a %o% b
```

<div class="alert alert-info">

**Vector "non-standard" operations**

Mathematically, vectors have no division. In R, `a / b` performs element-wise division.

```{r vec-div}
a <- 11:15
b <- a + 4

a / b
```

</div>

###

When two vectors have unequal length, R applies the **recycling rule**: the shorter vector is reused in cycles. With `e <- c(1, 2)` and `a` of length 5, `e` becomes `(1, 2, 1, 2, 1)`.

```{r cycle}
a <- 11:15
e <- c(1, 2)

a * e
```

R also issues a warning.

###

`c()` can build character vectors — wrap each element in double quotes:

```{r char-vec}
aSentence <- c("I", "am", "from", "People's Republic of China", ".")
aSentence
```


### Section Exercises

**Exercise 1.** Given the assignments below, explain and compute the result of each expression.

```{r q-recycle, exercise = TRUE}
a <- 11:15
e <- c(1, 2)

a - e
a / e
```

```{r q-recycle-solution}
# e is recycled to (1, 2, 1, 2, 1) to match a's length.
#
# a - e: 11-1, 12-2, 13-1, 14-2, 15-1 = 10 10 12 12 14
# a / e: 11/1, 12/2, 13/1, 14/2, 15/1 = 11 6 13 7 15
```

**Exercise 2.** Derive the sixty pairs of the Sexagenary Cycle.

China's traditional calendar pairs ten Heavenly Stems with twelve Earthly Branches. Of the 120 possible combinations, only sixty are canonical: Yang stems (odd positions) pair only with Yang branches; Yin stems (even positions) only with Yin branches.

- **Heavenly Stems:** 甲, 乙, 丙, 丁, 戊, 己, 庚, 辛, 壬, 癸
- **Earthly Branches:** 子, 丑, 寅, 卯, 辰, 巳, 午, 未, 申, 酉, 戌, 亥

> **Hints**
>
> - `seq()` with `%%` separates Yang from Yin.
> - `outer()` generates all pairwise combinations.
> - `paste0()` concatenates two strings (e.g. `paste0("甲", "子")` returns `"甲子"`).

```{r tiandi, exercise = TRUE}
tiangan <- c("甲", "乙", "丙", "丁", "戊", "己", "庚", "辛", "壬", "癸")
dizhi   <- c("子", "丑", "寅", "卯", "辰", "巳", "午", "未", "申", "酉", "戌", "亥")
```

```{r tiandi-solution}
tiangan <- c("甲", "乙", "丙", "丁", "戊", "己", "庚", "辛", "壬", "癸")
dizhi   <- c("子", "丑", "寅", "卯", "辰", "巳", "午", "未", "申", "酉", "戌", "亥")

tg_yang <- tiangan[seq(tiangan) %% 2 == 1]
tg_yin  <- tiangan[seq(tiangan) %% 2 == 0]

dz_yang <- dizhi[seq(dizhi) %% 2 == 1]
dz_yin  <- dizhi[seq(dizhi) %% 2 == 0]

outer(tg_yang, dz_yang, FUN = paste0)
outer(tg_yin,  dz_yin,  FUN = paste0)
```


## Matrices

Vectors are one-dimensional; matrices extend to two dimensions.

$$
A_{2 \times 2} = \begin{pmatrix} 1 & 3 \\ 2 & 4 \end{pmatrix}.
$$

```{r matrixA}
# 2-row matrix; data fill column by column
matrix(c(1, 2, 3, 4), nrow = 2)
```

`nrow` is "number of rows"; `ncol` is "number of columns". Specify only one — R computes the other. R fills matrices in **column-major order**. The recycling rule also applies during construction:

```{r matrix-recycle}
matrix(1:3, nrow = 2)
```

###

Elements are referenced by row and column. A blank position means "all".

```{r element}
A <- matrix(1:3, nrow = 2)

A[1, 2]
A[1, ]
A[, 2]
```


### Single-Matrix Operations

Transposition swaps rows and columns; in R, `t()`.

```{r matrixB}
A <- matrix(1:3, nrow = 2)

A
t(A)
t(t(A))
```

<div class="alert alert-warning">

**Nested function calls**

`t(t(A))` evaluates from the inside out: R first runs `t(A)`, then applies the outer `t()` to that result. This is identical in spirit to mathematical function composition $g(f(x))$.

</div>


### Matrix Operations

Matrix operations parallel their vector analogues. Addition and subtraction require matching dimensions. Multiplication has element-wise, inner, outer, and cross-product forms (`crossprod()` computes $X^{\!\top} X$ or $X^{\!\top} Y$).

```{r matrix_times}
B <- matrix(1:6, nrow = 2)
C <- matrix(6:1, nrow = 2)
D <- matrix(6:1, nrow = 3)

B
C
D

# Matching dimensions required
B + C
B - C

# Multiplication varieties
B * C            # element-wise (Hadamard)
B %*% D          # inner (dot) product
B %o% C          # outer (tensor) product
crossprod(B, C)  # cross-product: t(B) %*% C
```


### Solving Linear Systems

Many statistical methods reduce to solving linear systems. Consider:

$$
\begin{cases}
a + b + c = 6, \\
3a + 2b + 4c = 9, \\
2a + 2b - 6c = 4.
\end{cases}
$$

Letting **B** hold the coefficients, $\vec{y}$ the right-hand side, $\vec{x}$ the unknowns: $\vec{y} = \mathbf{B}\vec{x}$. This resembles a regression equation $y = \beta_0 + \beta_1 x_1 + \cdots + \beta_n x_n + \epsilon$. In R, `solve()` returns the answer:

```{r equations}
B <- matrix(c(1, 3, 2, 1, 2, 2, 1, 4, -6), nrow = 3)
B

y <- c(6, 9, 4)

x <- solve(B, y)
x
```


## Summary

- **Assignment:** `<-`
- **Boolean logic**: `==`, `!=`, `>`, `<`, `&`, `|`, `!`
- **Vectors**: `c()`, `seq()`, `[]`
- **Matrices**: `matrix()`, `nrow()`, `ncol()`, `t()`, `solve()`
