---
title: "Basic Mathematical Concepts"
subtitle: "Learn R with Dr. Hu and His Friends"
author: "Yue Hu (Department of Political Science, Tsinghua University)"
output:
  learnr::tutorial:
    language: "en"
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      out.width = "100%")
```

## Key Points

- Basic algebra
- Vectors
- Matrices and equations


## Basic Operations

The arithmetic symbols in R are identical to those in mathematics: addition (`+`), subtraction (`-`), multiplication (`*`), and division (`/`). In the R Console, type a formula as you would on paper, press Enter, and the result appears.

```{r algebra-basic}
# The four basic operations
1 + 2
3 - 4
5 * 6
7 / 8

# Compound expressions
(2 + 4) / (5 - 3)
```

<div class="alert alert-info">

**A note on whitespace**

You may have noticed that the formulas above place spaces between numbers and operators. This is purely a matter of readability — R itself is indifferent to whitespace. The expression `(2+4) /  (  5-   3        )` runs perfectly despite its irregular spacing.

For human readers, however, well-placed spaces help identify code units. Hadley Wickham's *Tidyverse Style Guide* recommends using spaces and blank lines to improve readability. This book follows that convention, and beginners are encouraged to develop the habit early. For inherited or third-party code, packages such as **styler** and **air** can reformat code in a single click.

</div>

###

Beyond the four basic operations, R supports a wider range of arithmetic: powers (`^`), modulo (`%%`), natural and base-10 logarithms (`log()`, `log10()`), and exponentials (`exp()`). Some operations are so common that they have dedicated functions named after their English abbreviations. Square root, for example, can be written as `^(1/2)`, but `sqrt()` ("square root") is more common. The same logic applies to `abs()` ("absolute value"). This naming convention dates back to FORTRAN and C.

```{r complex-math}
10^2
5 %% 3      # remainder
log(11)
log10(100)
exp(12)

# Two ways to take a square root
9^(1/2)
sqrt(9)

# Absolute value
abs(-9)
```

###

Other mathematical functions, though less frequently used, generally have R counterparts. Trigonometric functions are a case in point — note how the code reads almost like a passage from a math textbook:

```{r triangle}
sin(pi)
cos(pi)
tan(pi / 2)

acos(1)
asin(1)
atan(0.5)
```

###

In addition, R pre-defines a number of useful numerical and character constants, available for direct reuse. The constant `pi` above is one example, representing $\pi$. Other reserved words include the lowercase and uppercase Roman letters (`letters`, `LETTERS`) and month names and abbreviations (`month.name`, `month.abb`).

Try typing `letters` or `month.name` below and see what you get.

```{r reserved, exercise = TRUE}

```

```{r reserved-solution}
letters
month.name
```


### Scientific Notation

In the example above, `tan(pi / 2)` returned `1.633124e+16`. The expression of a number as the product of a real coefficient $a$ (with $1 \leq |a| < 10$) and an integer power of ten is known as **scientific notation**, attributed to Archimedes. In computing systems, the letter `e` ("exponential") connects the coefficient and the exponent: `e+16` means "times $10^{16}$". Thus $200 = 2 \times 10^2$ may be written as `2e2`, and $0.002 = 2 \times 10^{-3}$ as `2e-3`.

What does `2.8e+10` represent in plain numerals?

```{r sn, exercise = TRUE}

```

```{r sn-solution}
format(2.8e+10, scientific = FALSE, big.mark = ",")
```


### Boolean Algebra

Beyond arithmetic, R relies heavily on **Boolean algebra** — a logical algebraic system proposed by the Irish mathematician George Boole. In R, the two truth values are represented by `TRUE` and `FALSE` (uppercase), abbreviated as `T` and `F`. In numeric contexts these values are coerced automatically: `FALSE` becomes `0` and `TRUE` becomes `1` (so `TRUE + 1` evaluates to `2`). Three Boolean connectives are central: `&` (and), `|` (or), and `!` (not). The Boolean "empty set", `NULL`, is also a fixed object (uppercase).

Boolean algebra in R is most often paired with comparison operators: equality (`==`), inequality (`!=`), greater than (`>`), less than (`<`), greater than or equal to (`>=`), and less than or equal to (`<=`). Each example below returns either `TRUE` or `FALSE`.

```{r boolean}
# Equalities
1 == 2
1 == TRUE   # TRUE and FALSE behave as 1 and 0 with logical attributes
0 != F

# Inequalities
1 != 2
1 > 2
1 < 2
1 >= 2
1 <= 2

# Boolean operations
3 > 2 & 2 > 1
12 & 2 > 2
-3 | 2 > 2
```

###

Two points concerning the last three examples deserve attention. First, Boolean connectives are evaluated *after* their comparison operands: `3 > 2 & 2 > 1` first checks `3 > 2` and `2 > 1` (both `TRUE`), then computes `TRUE & TRUE`, which yields `TRUE`. Second, when a single value enters a Boolean expression — as in `12 & 2 > 2` and `-3 | 2 > 2` — it is treated as a self-equality check; that is, `12 & 2 > 2` is equivalent to `12 == 12 & 2 > 2`, and `-3 | 2 > 2` to `-3 == -3 | 2 > 2`.

<div class="alert alert-info">

**Why the double equal sign?**

Beginners often find it odd that "equality" requires two equal signs (`==`). One reason is that in many programming languages, including R and Python, the single equal sign `=` is reserved for **assignment** (we shall return to this in Chapter 4). Adding a second equal sign turns it into a comparison, keeping it consistent with `!=`, `>=`, and `<=`.

You may forget at first and write `=` when you mean `==` — there is no need to chastise yourself; every learner passes through this stage. Practice will resolve it.

</div>

What is the result of the following expression?

```{r judge, exercise = TRUE}
3 > 2 > 1
```

```{r judge-solution}
# In R, the expression evaluates left to right:
#   (3 > 2) -> TRUE -> 1
#   1 > 1  -> FALSE
# So the final result is FALSE — not the meaning that mathematical notation would suggest.
# To obtain the mathematical reading, write:
(3 > 2) & (2 > 1)
```


## Vectors

We now enter the realm of linear algebra, beginning with one of its foundational concepts: the **vector**. Mathematically, a vector is a directed sequence of numbers. The numbers from elementary algebra are, in linear-algebraic terms, **scalars** — quantities possessing magnitude but no direction.

In a math textbook, a vector takes the form $\vec{a} = (11, 12, 13, 14, 15)$. In R, the same vector is constructed with `a <- c(11, 12, 13, 14, 15)`, where `c()` stands for *concatenate* (or *combine*) — bundling elements into a single object. For consecutive integers the colon operator offers a shortcut: `a <- 11:15`. Mixed sequences are also possible, e.g. `a <- c(1, 3, 11:15, 20:25)`. A third route is the sequence function `seq()` (English: *sequence*): `a <- seq(from = 11, to = 15, by = 1)`. Because the step is one, this can be shortened to `a <- seq(11, 15)`.

```{r vector}
# Method 1: c() — the most general form
a  <- c(11, 12, 13, 14, 15)

# Method 2: the colon operator (consecutive integers only)
a1 <- 11:15

# Method 3: seq() — flexible step and length
a2 <- seq(from = 11, to = 15, by = 1)
a3 <- seq(11, 15)

# Verify that all four constructions yield identical results
a == a1
a == a2
a == a3
```

If you read `seq(from = 11, to = 15, by = 1)` aloud — "a sequence from 11 to 15 increased by 1" — that is precisely what it does. Many R commands read this way. Master a few patterns of expressing intent in plain English and the act of programming begins to feel less foreign.


### Indexing

In mathematical notation, an element of a vector is identified by a subscript: $\vec{a}_2 = 12$, $\vec{a}_5 = 15$, $\vec{a}_{3,4} = 13, 14$. In R, this is accomplished with square brackets `[]`, which extract a *subset* of the vector. The same mechanism becomes the foundation of data filtering in later chapters.

```{r indexing}
# Construct a demonstration vector
a <- c(11, 12, 13, 14, 15)

# Extract a single element by position
a[2]
a[5]

# Extract multiple elements via a position vector
a[c(3, 4)]
```

<div class="alert alert-info">

**The assignment arrow `<-`**

The arrow `<-` is the most common assignment operator in R. A useful mnemonic is to view it as a "pictograph": the arrow points left, indicating that the object on the right is *stored into* the name on the left. Once the vector `(11, 12, 13, 14, 15)` is saved as `a`, R thereafter regards `a` *as* that vector.

R is essentially the only popular language that uses `<-` for assignment; most others adopt the single equal sign. The reason traces back to S, R's predecessor (introduced in Chapter 2). In the 1960s, a notation system called **APL** (A Programming Language) developed into a highly expressive language and inspired many descendants, including S. APL employed many special characters, so a dedicated keyboard was designed; `<-` was the left-arrow key in the upper-right corner of that keyboard. APL faded, the keyboard layout did too, but the `<-` habit was preserved by developers and lives on in R.

The arrow makes data flow visually intuitive — source on the right, destination on the left. It has also become a sort of badge of honour among R enthusiasts: only `<-` users count as *true* R users. R provides a shortcut for typing it: **Alt + -** on Windows / Linux, **Option + -** on macOS.

</div>


### Vector Arithmetic

Vectors support addition, subtraction, and multiplication. The symbols (`+`, `-`, `*`) match those of elementary algebra. For addition and subtraction, corresponding elements are combined to yield a new vector:

$$\vec{a} \pm \vec{b} = (a_1 \pm b_1,\; a_2 \pm b_2,\; \ldots,\; a_n \pm b_n).$$

In R:

```{r vec-add}
a <- c(11, 12, 13, 14, 15)

# Vector ± scalar
b <- a + 4
c <- a - 0.5

# Vector ± vector
a + b
a - c
```

###

For multiplication, the asterisk `*` performs **element-wise multiplication**, also known as the *Hadamard product*:

$$\vec{a} \odot \vec{b} = a_{ij}\, b_{ij}.$$

A second, dimension-reducing form of multiplication condenses two vectors into a single number — the **inner product** (or *dot product*): $\vec{a} \cdot \vec{b} = \sum a_i b_i$. A special case is the scalar–vector product: $c\vec{a} = (ca_1, ca_2, \ldots, ca_n)$. In R, the inner product uses `%*%`.

The complementary operation, the **outer product** (or *tensor product*), expands two vectors into a matrix:

$$
\vec{a} \otimes \vec{b} = \mathbf{a}\mathbf{b}^{\!\top}.
$$

In R, the outer product uses `%o%` or the `outer()` function. Beyond pure arithmetic, the outer product can produce all pairwise combinations of two vectors — a feature we shall exploit shortly.

```{r vec-mult}
a <- c(11, 12, 13, 14, 15)
b <- a + 4

# Element-wise (Hadamard) product
a * b

# Inner (dot) product — returns a 1 × 1 matrix
a %*% b

# Outer (tensor) product — returns a matrix
a %o% b
```

<div class="alert alert-info">

**Vector "non-standard" operations**

Mathematically, neither dot nor outer product has a unique inverse, so there is no division of vectors in the strict sense. In R, however, `a / b` is well-defined: it performs element-wise division, computing `a[1]/b[1]`, `a[2]/b[2]`, and so forth.

</div>

```{r vec-div}
a <- 11:15
b <- a + 4

# Element-wise division
a / b
```

###

When two vectors have unequal length, R automatically applies the **recycling rule**: elements of the shorter vector are reused in cycles to match the longer one. For example, with `e <- c(1, 2)` and `a` of length 5, R extends `e` to `(1, 2, 1, 2, 1)` before performing the operation. The same applies to addition, subtraction, and division.

```{r cycle}
a <- 11:15
e <- c(1, 2)

a * e
```

R will also issue a warning, in effect saying: "you didn't give me enough — I had to recycle."

###

In addition, `c()` can construct **character vectors**; simply wrap each element in double quotes:

```{r char-vec}
# Each element is wrapped in double quotes
aSentence <- c("I", "am", "from", "People's Republic of China", ".")
aSentence
```


### Section Exercises

**Exercise 1.** Given the assignments below, explain and compute the result of each expression.

```{r q-recycle, exercise = TRUE}
a <- 11:15
e <- c(1, 2)

a - e
a * e
a / e
```

```{r q-recycle-solution}
# Vector a has length 5; vector e has length 2.
# R recycles e into (1, 2, 1, 2, 1) so that its length matches a.
#
# a - e:
#   11 - 1 = 10
#   12 - 2 = 10
#   13 - 1 = 12
#   14 - 2 = 12
#   15 - 1 = 14
# Result: 10 10 12 12 14
#
# a * e:
#   11*1, 12*2, 13*1, 14*2, 15*1 = 11, 24, 13, 28, 15
#
# a / e:
#   11/1, 12/2, 13/1, 14/2, 15/1 = 11, 6, 13, 7, 15
```

###

**Exercise 2.** *The Sexagenary Cycle (Heavenly Stems and Earthly Branches).*

China's traditional calendar uses a "Heavenly Stem + Earthly Branch" system, formed by combining the ten Heavenly Stems with the twelve Earthly Branches. There are $10 \times 12 = 120$ possible combinations in principle, but only sixty form the canonical cycle: by ancestral convention, **Yang stems pair only with Yang branches, and Yin stems with Yin branches** (where odd-positioned entries are Yang and even-positioned ones are Yin).

Use the operations introduced above to derive the sixty pairs in R.

- **Heavenly Stems:** 甲, 乙, 丙, 丁, 戊, 己, 庚, 辛, 壬, 癸
- **Earthly Branches:** 子, 丑, 寅, 卯, 辰, 巳, 午, 未, 申, 酉, 戌, 亥

```{r tiandi, exercise = TRUE}
tiangan <- c("甲", "乙", "丙", "丁", "戊", "己", "庚", "辛", "壬", "癸")
dizhi   <- c("子", "丑", "寅", "卯", "辰", "巳", "午", "未", "申", "酉", "戌", "亥")
```

> **Hints**
>
> - Use `seq()` together with the modulo operator `%%` to separate Yang from Yin.
> - `outer()` generates all pairwise combinations of two vectors.
> - To output the actual stem-branch text rather than positional indices, use `paste0()`, which concatenates two strings — for example, `paste0("甲", "子")` returns `"甲子"`.

```{r tiandi-solution}
tiangan <- c("甲", "乙", "丙", "丁", "戊", "己", "庚", "辛", "壬", "癸")
dizhi   <- c("子", "丑", "寅", "卯", "辰", "巳", "午", "未", "申", "酉", "戌", "亥")

# Step 1: split Yang (odd-indexed) and Yin (even-indexed) entries
tg_yang <- tiangan[seq(tiangan) %% 2 == 1]
tg_yin  <- tiangan[seq(tiangan) %% 2 == 0]

dz_yang <- dizhi[seq(dizhi) %% 2 == 1]
dz_yin  <- dizhi[seq(dizhi) %% 2 == 0]

# Step 2: outer product with paste0 to build all valid pairings
outer(tg_yang, dz_yang, FUN = paste0)
outer(tg_yin,  dz_yin,  FUN = paste0)
```


### Vector Properties

R provides built-in functions for summarising vectors. These will recur whenever we operate on data variables.

```{r vec-property}
a <- 11:15

# Length
length(a)

# Extrema
max(a)
min(a)

# Mean
mean(a)

# Median
median(a)

# Variance
var(a)
```


## Matrices

Vectors are one-dimensional — a single sequence of numbers. **Matrices** are the two-dimensional structures built from multiple vectors. In a math textbook, a matrix takes the form

$$
A_{2 \times 2} = \begin{pmatrix} 1 & 3 \\ 2 & 4 \end{pmatrix}.
$$

In R the same matrix appears as:

```{r matrixA}
# Create a 2-row matrix; data fill column by column
matrix(c(1, 2, 3, 4), nrow = 2)
```

The two are nearly indistinguishable — R even labels the row and column indices on your behalf.

The command `matrix(c(1, 2, 3, 4), nrow = 2)` reads in plain English: "make a *matrix* from the data `c(1, 2, 3, 4)` arranged in 2 rows." `nrow` is shorthand for "number of rows", a common R abbreviation. One may equivalently specify `ncol` (number of columns); only one of the two is required, and R computes the other automatically.

###

R fills matrices in **column-major order** — the first column first, then the second, and so on — consistent with FORTRAN but opposite to the row-first convention common in mathematical exposition. The **recycling rule** also applies during construction: if too few elements are supplied, R loops over the input until the matrix is full, accompanied by a warning.

```{r matrix-recycle}
# Asking for a 2-row matrix while supplying only three elements
matrix(1:3, nrow = 2)
```

###

Matrix elements are referenced by row and column subscripts. In mathematics, $A_{12}$ denotes row 1, column 2. The R equivalent is `A[1, 2]`. Leaving a position blank means "all": `A[1, ]` selects the entire first row, and `A[, 2]` selects the entire second column.

```{r element}
A <- matrix(1:3, nrow = 2)

# A specific element: row 1, column 2
A[1, 2]

# An entire row
A[1, ]

# An entire column
A[, 2]
```


### Single-Matrix Operations

The most common operation on a single matrix is **transposition** — interchanging its rows and columns. The transpose of $\mathbf{A}$ is conventionally denoted $\mathbf{A}^{\!\top}$. In R, transposition is performed by `t()`, named after the first letter of *transposition*.

```{r matrixB}
A <- matrix(1:3, nrow = 2)

# View the original matrix
A

# A single transpose
t(A)

# Two transpositions return the original
t(t(A))
```

<div class="alert alert-warning">

**Nested function calls**

The expression `t(t(A))` illustrates a common R idiom: nested function calls. The basic principle is *evaluate from the inside out*. R first runs `t(A)`, then feeds that result into the outer `t()`, repeating outwards until no further wrapping remains. The logic is identical to mathematical function composition such as $g(f(x))$.

</div>


### Matrix Operations

Matrix operations parallel their vector analogues:

- **Addition and subtraction** require matrices of the same dimensions: $A \pm B = a_{ij} \pm b_{ij}$.
- **Multiplication** comes in several flavours. The element-wise (Hadamard) product, inner product, and outer product all mirror their vector counterparts. For two matrices, the **inner product** (`%*%`) requires that the columns of the first matrix equal the rows of the second.
- R further provides `crossprod()`, which computes the **cross-product** $X^{\!\top} X$ or $X^{\!\top} Y$ — used for Gram and covariance matrices, and *not* identical to the mathematical outer product.

The example below uses three matrices:

- **B**: a 2-row matrix containing 1 to 6;
- **C**: a 2-row matrix containing 6 to 1;
- **D**: a 3-row matrix containing 6 to 1.

Addition and subtraction can be performed only between **B** and **C** (matching dimensions). Element-wise and outer products work analogously. The inner product requires *cols-of-first = rows-of-second*, so it can be performed only between **B** and **D**.

```{r matrix_times}
# Prepare matrices of different dimensions
B <- matrix(1:6, nrow = 2)
C <- matrix(6:1, nrow = 2)
D <- matrix(6:1, nrow = 3)

# Inspect their structure
B
C
D

# Basic operations (matching dimensions required)
B + C
B - C

# Multiplication exhibition: element-wise, inner, outer, cross
B * C            # element-wise (Hadamard)
B %*% D          # inner (dot) product
B %o% C          # outer (tensor) product
crossprod(B, C)  # cross-product: t(B) %*% C
```


### Solving Linear Systems

Why bother with matrix operations? Because **data frames** — the workhorse data structure of later chapters — are conceptually a special kind of matrix. Many operations defined for matrices (such as `t()`) extend to data frames, and many statistical methods reduce to linear-algebraic operations. Solving a multivariate equation, for instance, is mechanically very similar to estimating multivariate regression coefficients.

Consider the three-equation system

$$
\begin{cases}
a + b + c = 6, \\
3a + 2b + 4c = 9, \\
2a + 2b - 6c = 4.
\end{cases}
$$

Letting matrix **B** hold the coefficients, $\vec{y}$ the right-hand side, and $\vec{x}$ the unknowns, the system becomes

$$
\vec{y} = \mathbf{B} \vec{x}, \qquad
\begin{bmatrix} 6 \\ 9 \\ 4 \end{bmatrix} =
\begin{bmatrix} 1 & 1 & 1 \\ 3 & 2 & 4 \\ 2 & 2 & -6 \end{bmatrix}
\begin{bmatrix} b_1 \\ b_2 \\ b_3 \end{bmatrix}.
$$

This bears a striking resemblance to a regression equation:

$$
y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \cdots + \beta_n x_n + \epsilon.
$$

In R, `solve()` (English: *solve*) yields the solution in a single line:

```{r equations}
# Coefficient matrix B (3 × 3)
# Each column corresponds to the parameters of a, b, c
B <- matrix(c(1, 3, 2, 1, 2, 2, 1, 4, -6), nrow = 3)
B

# Constant vector y — the right-hand side: 6, 9, 4
y <- c(6, 9, 4)

# Solve the system: find x such that B %*% x = y
x <- solve(B, y)
x
```

Solving a system of equations is a direct stepping stone to fitting a regression model — a connection we shall return to in later chapters.


## Summary

This chapter introduces a number of concepts. Some need only an initial impression; their full meaning will deepen as later chapters revisit them in concrete settings. Others form the bedrock of what follows — failing to grasp them now will hinder later understanding.

The mathematical density of this chapter may feel heavy, but do not be intimidated. The aim is not to turn you into a mathematician; it is to help you understand the underlying logic by which R represents and processes data. R was designed for statistical computing, and statistical computing leans heavily on linear algebra. Vectors and matrices therefore become R's basic data containers, and Boolean, vector, and matrix operations together form the substrate of regression, clustering, and many other methods we shall meet later.

Four core modules to remember:

- **Assignment:** `<-`
- **Boolean logic**
    - Comparisons: `==` (equal), `!=` (not equal), `>`, `<`
    - Connectives: `&` (and), `|` (or), `!` (not)
- **Vectors and indexing**
    - Construction: `c()`
    - Sequence: `seq()`
    - Indexing: `[]`
- **Matrices**
    - Construction: `matrix()`
    - Dimensions: `nrow()`, `ncol()`
    - Transposition: `t()`
    - Solving / inversion: `solve()`
